---
title: "Air Pollution Interpolation & Clustering"
---

Load all required libraries
```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(sp)
library(rgdal)
library(tmap)
library(gstat)
library(spdep)
library(sf)
library(raster)
```


Loading the Monitor File
```{r}
Monitor_2020 <- read.csv("/Users/Shakeeb/Desktop/Air Pollution Interpolation & Clustering/annual_conc_by_monitor_2020.csv")
```

Our goal is to select 5 contiguous (compact states) in the United States to conduct spatial interpolation
and spatial clustering of annual Nitrogen Dioxide (NO2), Ozone (O3), and particulate matter 2.5 (PM2.5)
air pollution concentration.

We are given Illinois 2020, the other 4 contiguous states other than Illinois we will use are: Indiana, Iowa, Wisconsin and Missiouri.

State Codes: Illinois = 17, Indiana = 18, Iowa = 19, Missouri = 29, Wisconsin = 55, 
Nitrogen dioxide (NO2)
```{r}

IIIWM_NO2 <-  Monitor_2020 %>% filter(Monitor_2020$Parameter.Name == "Nitrogen dioxide (NO2)" & Monitor_2020$Pollutant.Standard == "NO2 Annual 1971" & (Monitor_2020$State.Code==17 | 
                                                                                    Monitor_2020$State.Code==18 | 
                                                                                    Monitor_2020$State.Code==19 | 
                                                                                    Monitor_2020$State.Code==29 | 
                                                                                    Monitor_2020$State.Code==55))


IIIWM_Ozone <-  Monitor_2020 %>% filter(Monitor_2020$Parameter.Name == "Ozone" & Monitor_2020$Pollutant.Standard == "Ozone 8-hour 2015" & (Monitor_2020$State.Code==17 | 
                                                                                    Monitor_2020$State.Code==18 | 
                                                                                    Monitor_2020$State.Code==19 | 
                                                                                    Monitor_2020$State.Code==29 | 
                                                                                    Monitor_2020$State.Code==55))


IIIWM_PM2.5 <-  Monitor_2020 %>% filter(Monitor_2020$Parameter.Name == "PM2.5 - Local Conditions" & Monitor_2020$Pollutant.Standard == "PM25 Annual 2012"  & (Monitor_2020$State.Code==17 | 
                                                                                    Monitor_2020$State.Code==18 | 
                                                                                    Monitor_2020$State.Code==19 | 
                                                                                    Monitor_2020$State.Code==29 | 
                                                                                    Monitor_2020$State.Code==55))

```


Remove remaining duplicates in data 
```{r}
final_NO2 <- IIIWM_NO2 %>% distinct(Latitude, Longitude, .keep_all = TRUE) 
final_Ozone <- IIIWM_Ozone %>% distinct(Latitude, Longitude, .keep_all = TRUE)
final_PM2.5 <- IIIWM_PM2.5 %>% distinct(Latitude, Longitude, .keep_all = TRUE)
```



Making the data into a spatial dataframes 
```{r}

proj_string <- sp::CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs")

NO2_cords <- final_NO2[, c("Longitude", "Latitude")]
spdf_NO2 <- sp::SpatialPointsDataFrame(coords = NO2_cords, data = final_NO2,
                                       proj4string = proj_string)

Ozone_cords <- final_Ozone[, c("Longitude", "Latitude")]
spdf_Ozone <- sp::SpatialPointsDataFrame(coords = Ozone_cords, data = final_Ozone,
                                       proj4string = proj_string)

PM2.5_cords <- final_PM2.5[, c("Longitude", "Latitude")]
spdf_PM2.5 <- sp::SpatialPointsDataFrame(coords = PM2.5_cords, data = final_PM2.5,
                                       proj4string = proj_string)

```


Add Study Area Boundary
```{r}

study_area <- readOGR("/Users/Shakeeb/Desktop/Air Pollution Interpolation & Clustering/Study Area", layer="Counties")

# Replace point boundary extent with that of Study Area
spdf_NO2@bbox <- study_area@bbox
spdf_Ozone@bbox <- study_area@bbox
spdf_PM2.5@bbox <- study_area@bbox

tm_shape(study_area) + tm_polygons() +
  tm_shape(spdf_NO2) +
  tm_dots(col="Arithmetic.Mean", palette = "-RdBu", auto.palette.mapping = FALSE,
             title="NO2 Average Annual Concentration", size=0.7) +
  tm_legend(legend.outside=TRUE)


tm_shape(study_area) + tm_polygons() +
  tm_shape(spdf_Ozone) +
  tm_dots(col="Arithmetic.Mean", palette = "-RdBu", auto.palette.mapping = FALSE,
             title="Ozone Average Annual Concentration", size=0.7) +
  tm_legend(legend.outside=TRUE)


tm_shape(study_area) + tm_polygons() +
  tm_shape(spdf_PM2.5) +
  tm_dots(col="Arithmetic.Mean", palette = "-RdBu", auto.palette.mapping = FALSE,
             title="PM2.5 Average Annual Concentration", size=0.7) +
  tm_legend(legend.outside=TRUE)
```



IDW and Leave One Out Cross Validation. Tested different k(idp) values for each pollutant to find best one 
```{r}

NO2IDW.out <- vector(length = length(spdf_NO2))
for (i in 1:length(spdf_NO2)) {
  NO2IDW.out[i] <- idw(Arithmetic.Mean ~ 1, spdf_NO2[-i,], spdf_NO2[i,], idp=4.0)$var1.pred
}
# Compute RMSE NO2
sqrt( sum((NO2IDW.out - spdf_NO2$Arithmetic.Mean)^2) / length(spdf_NO2))


OzoneIDW.out <- vector(length = length(spdf_Ozone))
for (i in 1:length(spdf_Ozone)) {
  OzoneIDW.out[i] <- idw(Arithmetic.Mean ~ 1, spdf_Ozone[-i,], spdf_Ozone[i,], idp=2.0)$var1.pred
}
# Compute RMSE Ozone
sqrt( sum((OzoneIDW.out - spdf_Ozone$Arithmetic.Mean)^2) / length(spdf_Ozone))


PM2.5IDW.out <- vector(length = length(spdf_PM2.5))
for (i in 1:length(spdf_PM2.5)) {
  PM2.5IDW.out[i] <- idw(Arithmetic.Mean ~ 1, spdf_PM2.5[-i,], spdf_PM2.5[i,], idp=3.0)$var1.pred
}
# Compute RMSE PM2.5
sqrt( sum((PM2.5IDW.out - spdf_PM2.5$Arithmetic.Mean)^2) / length(spdf_PM2.5))


```



Mapping IDW Interpolation
```{r}

# NO2- Create an empty grid where n is the total number of cells
grd_NO2              <- as.data.frame(spsample(spdf_NO2, "regular", n=50000))
names(grd_NO2)       <- c("X", "Y")
coordinates(grd_NO2) <- c("X", "Y")
gridded(grd_NO2)     <- TRUE  # Create SpatialPixel object
fullgrid(grd_NO2)    <- TRUE  # Create SpatialGrid object

proj4string(grd_NO2) <- proj4string(spdf_NO2)

NO2_idw_map <- gstat::idw(Arithmetic.Mean ~ 1, spdf_NO2, newdata=grd_NO2, idp=4.0)

NO2_raster <- raster(NO2_idw_map)
NO2_mask   <- mask(NO2_raster, study_area)

tm_shape(NO2_mask) + 
  tm_raster(n=10,palette = "-RdBu", auto.palette.mapping = FALSE,
            title="NO2 Average Annual Concentration \n (K=4)") + 
  tm_shape(study_area) + tm_borders(col = "black") +
  tm_shape(spdf_NO2) + tm_dots(size=0.2) +
  tm_legend(legend.outside=TRUE)


# Ozone - Create an empty grid where n is the total number of cells
grd_Ozone              <- as.data.frame(spsample(spdf_Ozone, "regular", n=50000))
names(grd_Ozone)       <- c("X", "Y")
coordinates(grd_Ozone) <- c("X", "Y")
gridded(grd_Ozone)     <- TRUE  # Create SpatialPixel object
fullgrid(grd_Ozone)    <- TRUE  # Create SpatialGrid object

proj4string(grd_Ozone) <- proj4string(spdf_Ozone)

Ozone_idw_map <- gstat::idw(Arithmetic.Mean ~ 1, spdf_Ozone, newdata=grd_Ozone, idp=2.0)

Ozone_raster <- raster(Ozone_idw_map)
Ozone_mask   <- mask(Ozone_raster, study_area)

tm_shape(Ozone_mask) + 
  tm_raster(n=10,palette = "-RdBu", auto.palette.mapping = FALSE,
            title="Ozone Average Annual Concentration \n (K=2)") +
  tm_shape(study_area) + tm_borders(col = "black") +
  tm_shape(spdf_Ozone) + tm_dots(size=0.2) +
  tm_legend(legend.outside=TRUE)


# PM2.5 - Create an empty grid where n is the total number of cells
grd_PM2.5              <- as.data.frame(spsample(spdf_PM2.5, "regular", n=50000))
names(grd_PM2.5)       <- c("X", "Y")
coordinates(grd_PM2.5) <- c("X", "Y")
gridded(grd_PM2.5)     <- TRUE  # Create SpatialPixel object
fullgrid(grd_PM2.5)    <- TRUE  # Create SpatialGrid object

proj4string(grd_PM2.5) <- proj4string(spdf_PM2.5)

PM2.5_idw_map <- gstat::idw(Arithmetic.Mean ~ 1, spdf_PM2.5, newdata=grd_PM2.5, idp=3.0)

PM2.5_raster <- raster(PM2.5_idw_map)
PM2.5_mask   <- mask(PM2.5_raster, study_area)

tm_shape(PM2.5_mask) + 
  tm_raster(n=10,palette = "-RdBu", auto.palette.mapping = FALSE,
            title="PM2.5 Average Annual Concentration \n (K=3)") +
  tm_shape(study_area) + tm_borders(col = "black") +
  tm_shape(spdf_PM2.5) + tm_dots(size=0.2) +
  tm_legend(legend.outside=TRUE) 

```



Kriging


Plotting air pollution concentrations
```{r}

ggplot(data = final_NO2) + 
  geom_point(mapping = aes(x = Longitude, y = Latitude, size = Arithmetic.Mean),
             color="brown", alpha=3/4) + ggtitle("NO2 Average Annual Concentration")

ggplot(data = final_Ozone) + 
  geom_point(mapping = aes(x = Longitude, y = Latitude, size = Arithmetic.Mean),
             color="brown", alpha=3/4) + ggtitle("Ozone Average Annual Concentration")

ggplot(data = final_PM2.5) + 
  geom_point(mapping = aes(x = Longitude, y = Latitude, size = Arithmetic.Mean),
             color="brown", alpha=3/4) + ggtitle("PM2.5 Average Annual Concentration")
```



# Fitting a Variogram

First need to check if variables are normally distributed
```{r}
hist(spdf_NO2$Arithmetic.Mean)
hist(spdf_Ozone$Arithmetic.Mean)
hist(spdf_PM2.5$Arithmetic.Mean)
shapiro.test(spdf_NO2$Arithmetic.Mean)
shapiro.test(spdf_Ozone$Arithmetic.Mean)
shapiro.test(spdf_PM2.5$Arithmetic.Mean)

#Since they are we can continue to fitting the variogram
```


# Calculate Sample Variograms
```{r}
NO2_vgm <- variogram(Arithmetic.Mean~1, spdf_NO2)
plot(NO2_vgm)




Ozone_vgm <- variogram(Arithmetic.Mean~1, spdf_Ozone)
plot(Ozone_vgm)




PM2.5_vgm <- variogram(Arithmetic.Mean~1, spdf_PM2.5)
plot(PM2.5_vgm)




```

View different models that can be used to fit data
```{r}
show.vgms()
```
Fit the variogram
```{r}
# We test multiple model types for each pollutant and then optimize sill, nugget, and range

NO2_fit <- fit.variogram(NO2_vgm, model = vgm(psill = 25, nugget = 5, range = 100, model = "Mat"))
plot(NO2_vgm, NO2_fit)


Ozone_fit <- fit.variogram(Ozone_vgm, model = vgm(psill = 5.1e-06, nugget =  4.5e-06, range = 230, 
                                                  model = "Lin"))
plot(Ozone_vgm, Ozone_fit)


PM2.5_fit <- fit.variogram(PM2.5_vgm, model = vgm(psill = 1, nugget = 0.7, range = 180, model = "Gau"))
plot(PM2.5_vgm, PM2.5_fit)
attr(NO2_fit, "SSErr")
attr(Ozone_fit, "SSErr")
attr(PM2.5_fit, "SSErr")
```


## Cross-Validation 

Use LOOCV
```{r}
# We check the RMSE of the different model types to see which one fits the best

RMSE_resid <- function(x){
  return(sqrt(mean(x^2)))
} 

NO2_LOOCV_krige <-krige.cv(Arithmetic.Mean~1, spdf_NO2, model=NO2_fit)
RMSE_resid(NO2_LOOCV_krige@data$residual)


Ozone_LOOCV_krige <-krige.cv(Arithmetic.Mean~1, spdf_Ozone, model=Ozone_fit)
RMSE_resid(Ozone_LOOCV_krige@data$residual)


PM2.5_LOOCV_krige <-krige.cv(Arithmetic.Mean~1, spdf_PM2.5, model=PM2.5_fit)
RMSE_resid(PM2.5_LOOCV_krige@data$residual)
```



Kriging Prediction & Map Creation
```{r}

NO2_krg <- krige(Arithmetic.Mean~1, spdf_NO2, grd_NO2, NO2_fit)
Ozone_krg <- krige(Arithmetic.Mean~1, spdf_Ozone, grd_Ozone, Ozone_fit)
PM2.5_krg <- krige(Arithmetic.Mean~1, spdf_PM2.5, grd_PM2.5, PM2.5_fit)

NO2_r <- raster(NO2_krg)
NO2_m <- mask(NO2_r, study_area)
tm_shape(NO2_m) + 
  tm_raster(n=10, palette="-RdBu", auto.palette.mapping=FALSE, 
            title="NO2 Average Annual Concentration \n(Model=Mat)") +
  tm_shape(study_area) + tm_borders(col = "black") +
  tm_shape(spdf_NO2) + tm_dots(size=0.2) +
  tm_legend(legend.outside=TRUE)


Ozone_r <- raster(Ozone_krg)
Ozone_m <- mask(Ozone_r, study_area)
tm_shape(Ozone_m) + 
  tm_raster(n=10, palette="-RdBu", auto.palette.mapping=FALSE, 
            title="Ozone Average Annual Concentration \n(Model=Lin)") +
  tm_shape(study_area) + tm_borders(col = "black") +
  tm_shape(spdf_Ozone) + tm_dots(size=0.2) +
  tm_legend(legend.outside=TRUE)



PM2.5_r <- raster(PM2.5_krg)
PM2.5_m <- mask(PM2.5_r, study_area)
tm_shape(PM2.5_m) + 
  tm_raster(n=10, palette="-RdBu", auto.palette.mapping=FALSE, 
            title="PM2.5 Average Annual Concentration \n(Model=Gau)") +
  tm_shape(study_area) + tm_borders(col = "black") +
  tm_shape(spdf_PM2.5) + tm_dots(size=0.2) +
  tm_legend(legend.outside=TRUE)
```





## extracting the data for Clustering

```{r}

study_area$NO2mean <- raster::extract(NO2_raster,study_area,fun=mean)
study_area$PMmean <- raster::extract(PM2.5_raster,study_area,fun=mean)
study_area$Ozonemean <- raster::extract(Ozone_raster,study_area,fun=mean)
vars <- c("NO2mean","PMmean","Ozonemean")
study_area_dat <- data.frame(study_area@data[,vars])

sdat <- scale(study_area_dat)

```

## Neighbor list
```{r}
study_area.nb <- poly2nb(study_area)

plot(study_area, border=grey(.5))
plot(study_area.nb, coordinates(study_area), col="blue", add=TRUE)
```

## Minimum spanning tree
```{r}

lcosts <- nbcosts(study_area.nb,sdat)

study_area.w <- nb2listw(study_area.nb,lcosts,style="B")

study_area.mst <- mstree(study_area.w)


plot(study_area.mst,coordinates(study_area),col="blue",cex.lab=0.7)
plot(study_area,border=gray(.5),add=TRUE)
```


## Contiguity-constrained cluster

```{r}
clus4 <- skater(study_area.mst[,1:2],sdat,2)
str(clus4)

plot(clus4,coordinates(study_area),cex.lab=0.7,
    groups.colors=c("red","green","blue"))
plot(study_area,border=gray(.5),add=TRUE)


plot(study_area,col=c("red","green","blue")[clus4$groups],main = "Spatial Clustering for 3 Pollutents in the Study Area")
legend(x = "topright",legend = c("NO2","PM2.5","Ozone"),fill =c("red","green","blue"))

```


